---
source: Rmd
title: Data visualization
teaching: 60
exercises: 30
---

```{r loaddata, echo=FALSE, purl=FALSE, message=FALSE}
if (!file.exists("data/rnaseq.csv"))
  download.file(url = "https://github.com/example/data/rnaseq.csv",
                destfile = "data/rnaseq.csv")
```

::::::::::::::::::::::::::::::::::::::: objectives

- Produce scatter plots and boxplots using ggplot.
- Modify the aesthetics of an existing ggplot plot (including axis labels and color).

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How can I create publication-quality plots in R?

::::::::::::::::::::::::::::::::::::::::::::::::::

## Data Visualization

We start by loading the required packages. **`ggplot2`** is included in
the **`tidyverse`** package.

```{r load-package, message=FALSE, purl=TRUE}
library("tidyverse")
```

If not still in the workspace, load the data we saved in the previous lesson.

```{r load-data, eval=FALSE, purl=TRUE}
rna <- read.csv("data/rnaseq.csv")
```

The [Data Visualization Cheat Sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf)
covers the basics of `ggplot2`. The `ggplot2` webpage
([https://ggplot2.tidyverse.org](https://ggplot2.tidyverse.org)) provides ample documentation.

## Plotting with `ggplot2`

`ggplot2` is a plotting package that makes it simple to create complex
plots from data in a data frame. The theoretical foundation is the
*Grammar of Graphics* (@Wilkinson:2005).

To build a ggplot, use the following basic template:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) + <GEOM_FUNCTION>()
```

- use the `ggplot()` function and bind the plot to a specific data frame

```{r first-ggplot, cache=FALSE, purl=TRUE, fig.alt="Default histogram produced by ggplot() and geom_histogram() for the expression data."}
ggplot(data = rna, mapping = aes(x = expression)) +
  geom_histogram()
```

[^note]: Source: [Data Visualization Cheat Sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf).

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge

You have probably noticed an automatic message that appears when
drawing the histogram. Change the argument `bins` of `geom_histogram()`
to change the number of bins.

:::::::::::::::  solution

## Solution

```{r solution-bins, purl=TRUE, fig.alt="Histogram with 15 bins produced by ggplot() and geom_histogram()."}
ggplot(rna, aes(x = expression)) +
    geom_histogram(bins = 15)
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

We can apply a log2 transformation to have a more symmetric distribution.
Note the small constant (`+1`) to avoid `-Inf` values for zero counts [^note].

```{r log-transfo, cache=FALSE, purl=TRUE, fig.alt="Histogram of log2-transformed expression values showing a more symmetric distribution."}
rna <- rna |>
  mutate(expression_log = log2(expression + 1))
```

See also the video tutorials ([part 1](https://www.youtube.com/watch?v=h29g21z0a68) and
[part 2](https://www.youtube.com/watch?v=0m4yywqNPVY)) by Thomas Lin Pedersen.

:::::::::::::::::::::::::::::::::::::::: keypoints

- Use `ggplot2` for publication-quality visualizations in R.
- Add layers with the `+` operator to build complex plots step by step.

::::::::::::::::::::::::::::::::::::::::::::::::::
